---
title: "Dynamic QSP reporting with R and RMarkdown"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

_This is an example of creation of a dynamic QSP report in R and RMarkdown build based on heta-compiler and mrgsolve. The content of this file and modeling platform is published in GitHub repository <https://github.com/insysbio/insulin-signaling-t2d>_

## Preamble

The QSP model which is used as an example of QSP model was published in the article:

> Brannmark C, Nyman E, Fagerholm S, Bergenholm L, Ekstrand EM, Cedersund G, Stralfors P. Insulin Signaling in Type 2 Diabetes: Experimental and modeling analyses reveal mechanisms of insulin resistance in human adipocytes. Journal of biological chemistry.. 2013 288(14):9867–9880. DOI: 10.1074/jbc.M112.432062

The SBML code was downloaded from BioModels <https://www.ebi.ac.uk/biomodels/BIOMD0000000448> and used as the part of the [Heta-based](https://hetalang.github.io/#/) modeling platform.

The report includes the steps to reproduce simulations from the original article demonstration facilities of the approach and necessary setups.

All necessary files can also be found in the [repository](https://github.com/insysbio/insulin-signaling-t2d).

## Preparations

For easier creation of the Heta-based platform install [heta compiler](https://hetalang.github.io/#/heta-compiler/?id=installation).

In command line interface run the code below to create heta platform template

```sh
heta init
```
The minimal content will be created.

Download the SBML model [from the database](https://www.ebi.ac.uk/biomodels/BIOMD0000000448) and copy it into `src/BIOMD0000000448.xml`

Update the `src/index.heta` with the following content:

```heta
// load SBML model as a content of the platform
include BIOMD0000000448_url.xml type sbml

// make the records of a model observable
block {output: true} begin
  measuredIRp;
  measuredIRint;
  measuredIRS1p;
  measuredIRS1307;
  measuredPKB308p;
  measuredPKB473p;
  measuredAS160p;
  measuredmTORC1a;
  measuredS6Kp;
  glucoseuptake;
  measuredmTORC2a;
  measuredS6p;
end

// make insulin as an input of the model
insulin @Const = 10; // nM

// make IR (insuline resistance) specific parameters as an input of the model
IR_total @Const = 100;    // reduce to 55%
GLUT4_total @Const = 100; // GLUT4 reduce to 50%
diabetes @Const = 1;      // reduce to 15%

// recalculate initial values for IR and base conditions
IR .= IR_total * 99.87/100; // 99.87
IRp .= 0;
IRins .= 0;
IRip .= IR_total * 0.02/100;      // 0.02
IRi .= IR_total * 0.11/100;       // 0.11
//
GLUT4 .= GLUT4_total * 73.48/100;   // 73.48
GLUT4m .= GLUT4_total * 26.52/100;  // 26.52

// variable parameters
k1a @Const = 0.6331;
k1basal @Const = 0.03683;
k1c @Const = 0.8768;
k1d @Const = 31.01;

#export { format: Mrgsolve, filepath: _mrgsolve };
```

Install [R](https://www.r-project.org/) and add all necessary packages in Julia's console:

```{r eval=FALSE}
install.packages('mrgsolve')
install.packages('ggplot2')
```

## Loading platform

```{r message=FALSE, warning=FALSE}
library('mrgsolve')
library('ggplot2')
```

```{r}
system('heta build --dist-dir . ..', intern = TRUE)
```

```{r}
m <- mrgsolve::mread(model = 'nameless', file = '../_mrgsolve/nameless.cpp')

observables <- paste(
  'measuredIRp',
  'measuredIRint',
  'measuredIRS1p',
  'measuredIRS1307',
  'measuredPKB308p',
  'measuredPKB473p',
  'measuredAS160p',
  'measuredmTORC1a',
  'measuredS6Kp',
  'glucoseuptake',
  'measuredmTORC2a',
  'measuredS6p',
  sep=', '
)
```

## Default simulation

```{r}
m %>% mrgsim(end = 30, outvars = observables) %>% plot

m %>% param(IR_total=55, GLUT4_total=50, diabetes=0.15) %>% mrgsim(end = 30, outvars = observables) %>% plot
```

## Simulation scenarios

```{r}
scn_base <- m %>% update(end = 30, outvars = observables)
scn_ir <- m %>% param(IR_total=55, GLUT4_total=50, diabetes=0.15) %>% update(end = 30, outvars = observables)
```

## Advanced visualization

```{r}
results_df_base <- mrgsim(scn_base) %>% as.data.frame
results_df_ir <- mrgsim(scn_ir) %>% as.data.frame


results_df_base$scenario <- 'base'
results_df_ir$scenario <- 'ir'
results_df <- rbind(results_df_base, results_df_ir)

knitr::kable(head(results_df), format="html")
```

```{r}
ggplot(results_df, aes(x=time, y=measuredIRp)) +
  geom_line(aes(col=scenario)) +
  labs(title="Fig5 1B")
```

```{r}
ggplot(results_df, aes(x=time, y=measuredIRint)) +
  geom_line(aes(col=scenario)) +
  labs(title="Fig5 1C")
```
```{r}
ggplot(results_df, aes(x=time, y=measuredIRS1p)) +
  geom_line(aes(col=scenario)) +
  labs(title="Fig5 2B")
```
```{r}
ggplot(results_df, aes(x=time, y=measuredIRS1p)) +
  geom_line(aes(col=scenario)) +
  labs(title="Fig5 2C")
```
```{r}
ggplot(results_df, aes(x=time, y=measuredPKB308p)) +
  geom_line(aes(col=scenario)) +
  labs(title="Fig5 3B")
```
```{r}
ggplot(results_df, aes(x=time, y=measuredPKB473p)) +
  geom_line(aes(col=scenario)) +
  labs(title="Fig5 3C")
```

