---
title: "Dynamic QSP reporting with R and RMarkdown"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

_This is an example of creation of a dynamic QSP report in R and RMarkdown build based on heta-compiler and mrgsolve. 
The content of this file and modeling platform is published in GitHub repository <https://github.com/insysbio/insulin-signaling-t2d>_

## Preamble

The QSP model which is used as an example of QSP model was published in the article:

> Brannmark C, Nyman E, Fagerholm S, Bergenholm L, Ekstrand EM, Cedersund G, Stralfors P. Insulin Signaling in Type 2 Diabetes: Experimental and modeling analyses reveal mechanisms of insulin resistance in human adipocytes. Journal of biological chemistry.. 2013 288(14):9867–9880. DOI: 10.1074/jbc.M112.432062

The SBML code was downloaded from BioModels <https://www.ebi.ac.uk/biomodels/BIOMD0000000448> and used as the part of the [Heta-based](https://hetalang.github.io/#/) modeling platform.

The report includes the steps to reproduce simulations from the original article demonstration facilities of the approach and necessary setups.

All necessary files can also be found in the [repository](https://github.com/insysbio/insulin-signaling-t2d).

## Preparations

For easier creation of the Heta-based platform install [heta compiler](https://hetalang.github.io/#/heta-compiler/?id=installation).

In command line interface (console) run the code below to create heta platform template

```sh
heta init
```
The minimal content will be created.

Download the SBML model [from the database](https://www.ebi.ac.uk/biomodels/BIOMD0000000448) and copy it into `src/BIOMD0000000448.xml`

Update the `src/index.heta` with the following content:

```heta
// load SBML model as a content of the platform
include BIOMD0000000448_url.xml type sbml

// make the records of a model observable
block {output: true} begin
  measuredIRp;
  measuredIRint;
  measuredIRS1p;
  measuredIRS1307;
  measuredPKB308p;
  measuredPKB473p;
  measuredAS160p;
  measuredmTORC1a;
  measuredS6Kp;
  glucoseuptake;
  measuredmTORC2a;
  measuredS6p;
end

// make insulin to be an input of the model
insulin @Const = 10; // nM

// make IR (insuline resistance) specific parameters to be input of the model
IR_total @Const = 100;    // reduce to 55%
GLUT4_total @Const = 100; // GLUT4 reduce to 50%
diabetes @Const = 1;      // reduce to 15%

// recalculate initial values for IR and base conditions
IR .= IR_total * 99.87/100; // 99.87
IRp .= 0;
IRins .= 0;
IRip .= IR_total * 0.02/100;      // 0.02
IRi .= IR_total * 0.11/100;       // 0.11
//
GLUT4 .= GLUT4_total * 73.48/100;   // 73.48
GLUT4m .= GLUT4_total * 26.52/100;  // 26.52

// variable parameters
k1a @Const = 0.6331;
k1basal @Const = 0.03683;
k1c @Const = 0.8768;
k1d @Const = 31.01;

#export { format: Mrgsolve, filepath: _mrgsolve };
```

Install [R](https://www.r-project.org/), [RStudio](https://www.rstudio.com/products/rstudio/) and add all necessary packages:

```{r eval=FALSE}
install.packages('mrgsolve')
install.packages('ggplot2')
```

## Loading platform

```{r message=FALSE, warning=FALSE}
library('mrgsolve')
library('ggplot2')
```

The approach uses __mrgsolve__ package as a simulation engine and __Heta compiler__ is a connector to SBML and Heta formats.
Heta compiler can be run from the R environment. Mind the current working directory and the location of the Heta-based platform: `..`.

```{r}
system('heta build --dist-dir . ..', intern = TRUE)
```

After compiling model it can be loaded into mrgsolve.

```{r}
m <- mrgsolve::mread(model = 'nameless', file = '../_mrgsolve/nameless.cpp')

# list of observables
observables <- paste(
  'measuredIRp',
  'measuredIRint',
  'measuredIRS1p',
  'measuredIRS1307',
  'measuredPKB308p',
  'measuredPKB473p',
  'measuredAS160p',
  'measuredmTORC1a',
  'measuredS6Kp',
  'glucoseuptake',
  'measuredmTORC2a',
  'measuredS6p',
  sep=', '
)
```

## Default simulation

To simulate the observables only one line of code is required. Here we are using "chain" with `%>%` syntax.

```{r}
m %>% mrgsim(end = 30, outvars = observables) %>% plot
```

Based on the original publication the insuline resistance (IR) can be set by updating three parameters.

```{r}
m %>% param(IR_total=55, GLUT4_total=50, diabetes=0.15) %>% mrgsim(end = 30, outvars = observables) %>% plot
```

## Simulation scenarios

Mrgsolve does not operate with "scenario/condition" terms.
This functionality can be reproduced by `update` function.

```{r}
scn_base <- m %>% update(end = 30, outvars = observables)
scn_ir <- m %>% param(IR_total=55, GLUT4_total=50, diabetes=0.15) %>% update(end = 30, outvars = observables)
```

## Advanced visualization

If you need to reshape and visualize different simulation results the easier way is to convert simulation results into `data.frame`.

The following simulations reproduce the figures from the original paper.

```{r}
results_df_base <- mrgsim(scn_base) %>% as.data.frame
results_df_ir <- mrgsim(scn_ir) %>% as.data.frame


results_df_base$scenario <- 'base'
results_df_ir$scenario <- 'ir'
results_df <- rbind(results_df_base, results_df_ir)

knitr::kable(head(results_df), format = "html")
```

```{r}
ggplot(results_df, aes(x = time, y = measuredIRp)) +
  geom_line(aes(col = scenario)) +
  labs(title = "Fig5 1B")
```

```{r}
ggplot(results_df, aes(x = time, y = measuredIRint)) +
  geom_line(aes(col = scenario)) +
  labs(title = "Fig5 1C")
```
```{r}
ggplot(results_df, aes(x = time, y = measuredIRS1p)) +
  geom_line(aes(col = scenario)) +
  labs(title = "Fig5 2B")
```
```{r}
ggplot(results_df, aes(x = time, y = measuredIRS1p)) +
  geom_line(aes(col = scenario)) +
  labs(title = "Fig5 2C")
```
```{r}
ggplot(results_df, aes(x = time, y = measuredPKB308p)) +
  geom_line(aes(col = scenario)) +
  labs(title = "Fig5 3B")
```
```{r}
ggplot(results_df, aes(x = time, y = measuredPKB473p)) +
  geom_line(aes(col = scenario)) +
  labs(title = "Fig5 3C")
```

## Titration-like simulations

The simulations of another type (not time dependence) can be performed and visualized by applying `idata` approach. 
For example in the original article the titration-like experiment is simulated: 
intake of insulin and measurements of different observables after 10 minutes.

To reproduce the figures the following `idata` table should be created:

```{r}
titr_scn_df <- data.frame(ID = 1:22, insulin=c(
  1.00E-03,
  3.16E-03,
  1.00E-02,
  3.16E-02,
  1.00E-01,
  3.16E-01,
  1.00E+00,
  3.16E+00,
  1.00E+01,
  3.16E+01,
  1.00E+02,
  1.00E-03,
  3.16E-03,
  1.00E-02,
  3.16E-02,
  1.00E-01,
  3.16E-01,
  1.00E+00,
  3.16E+00,
  1.00E+01,
  3.16E+01,
  1.00E+02
))
```

```{r}
results_titr_base <- scn_base %>%
  idata_set(titr_scn_df) %>%
  mrgsim(end = 10) %>%
  filter(time == 10)
results_titr_ir <- scn_ir %>%
  idata_set(titr_scn_df) %>%
  mrgsim(end = 10) %>%
  filter(time == 10)

results_titr_base$scenario <- 'base'
results_titr_ir$scenario <- 'ir'
results_titr <- rbind(results_titr_base, results_titr_ir)

results_titr$insulin <- titr_scn_df$insulin[match(results_titr$ID, titr_scn_df$ID)]
```

```{r}
ggplot(results_titr, aes(x=insulin, y=measuredmTORC2a)) +
  geom_line(aes(col=scenario)) +
  scale_x_log10() +
  labs(title="Fig5 1A")
```

## Multiple simulations

`idata` can also be used to run Monte-Carlo simulations based on parameter variability.
For the demonstration purposes we will generate a random set of parameters: k1a, k1basal, k1c, k1d.
This simulations mimic the uncertainty in the selected parameters.

```{r}
mc_scn_df <- data.frame(
  ID = 1:100,
  k1a = rlnorm(100, mean = 0.6331, sd = 0.5),
  k1basal = rlnorm(100, mean = 0.03683, sd = 0.5),
  k1c = rlnorm(100, mean = 0.8768, sd = 0.5),
  k1d = rlnorm(100, mean = 31.01, sd = 0.5)
)
```

```{r}
results_mc_base <- scn_base %>%
  idata_set(mc_scn_df) %>%
  mrgsim(outvars = 'measuredIRp, measuredIRint')
results_mc_ir <- scn_ir %>%
  idata_set(mc_scn_df) %>%
  mrgsim(outvars = 'measuredIRp, measuredIRint')
```

```{r}
results_mc_base %>% plot
results_mc_ir %>% plot
```
